Analysis

### 2022-07-04 CTCF_Threshold_mm.Rmd
Latest changes

```{r settings}
# Project folder path
dir_data <- "/Users/mdozmorov/Documents/Work/GitHub/CTCF.dev"

# Input files
fileNameIn1 <- "/Users/mdozmorov/Documents/Data/GoogleDrive/CTCF.dev/merlot/hg38.JASPAR2022_MA0139.1.bed"
fileNameIn2 <- file.path(dir_data, "data/GRCh38-CTCF.bed")
# Output files
fileNameOut1 <- file.path(dir_data, "")
```

@@

fileNameOut1 <- file.path(dir_data, "results", "Figure_human_pvalues_threshold.svg")
ggsave(filename = fileNameOut1, width = 4.5, height = 1.5)
# dev.off()
# svg(filename = fileNameOut1, width = 7, height = 3)
# print(p)
# dev.off()

>>>>

```{r settings}
dir_data <- "/Users/mdozmorov/Documents/Data/GoogleDrive/CTCF.dev"
# Project folder path
dir_project <- "/Users/mdozmorov/Documents/Work/GitHub/CTCF.dev"

# Input files
fileNameIn1 <- file.path(dir_project, "RData", "mm10.MA0139.1.bed")
fileNameIn2 <- file.path(dir_project, "data", "mm10-CTCF.bed")
# Output files
fileNameOut1 <- file.path(dir_project, "results", "Figure_mouse_pvalues_threshold.svg")
```

@@

ggsave(filename = fileNameOut1, width = 4.5, height = 1.5)
# dev.off()
# svg(filename = fileNameOut1, width = 7, height = 3)
# print(p)
# dev.off()

### 2022-07-04 EDA_SCREEN.Rmd
Latest changes

```{r settings}
# Project folder path
dir_data <- "/Users/mdozmorov/Documents/Work/GitHub/CTCF.dev"

# Input files
# Human CTCF-bound cCRE (hg38)
fileNameIn1 <- file.path(dir_data, "data/GRCh38-CTCF.bed")
# download.file("https://api.wenglab.org/screen_v13/fdownloads/cCREs/GRCh38-CTCF.bed", fileNameIn1)
# Mouse CTCF-bound cCRE (mm10)
fileNameIn2 <- file.path(dir_data, "data/mm10-CTCF.bed")
# download.file("https://api.wenglab.org/screen_v13/fdownloads/cCREs/mm10-CTCF.bed", fileNameIn2)
# Output files
fileNameOut1 <- file.path(dir_data, "results/CTCF_hg38_SCREEN.RData")
fileNameOut2 <- file.path(dir_data, "results/CTCF_mm10_SCREEN.RData")
```

@@

  return(gr)
}

# Save as RData object. subdir is the character name of the CTCF GRanges variable
save(list = mtx_to_gr(mtx = mtx_human, genome_id = "hg38"), file = fileNameOut1)
save(list = mtx_to_gr(mtx = mtx_mouse, genome_id = "mm10"), file = fileNameOut2)
# load(file = paste0(subdir, ".RData"))
```

>>>>

```{r settings}
# Project folder path
dir_project <- "/Users/mdozmorov/Documents/Work/GitHub/CTCF.dev"

# Input files
# Human CTCF-bound cCRE (hg38)
fileNameIn1 <- file.path(dir_project, "data/GRCh38-CTCF.bed")
if (!file.exists(fileNameIn1)) {
  download.file("https://api.wenglab.org/screen_v13/fdownloads/cCREs/GRCh38-CTCF.bed", fileNameIn1)
}
# Mouse CTCF-bound cCRE (mm10)
fileNameIn2 <- file.path(dir_project, "data/mm10-CTCF.bed")
if (!file.exists(fileNameIn2)) {
  download.file("https://api.wenglab.org/screen_v13/fdownloads/cCREs/mm10-CTCF.bed", fileNameIn2)
}
# Output files
fileNameOut1 <- file.path(dir_project, "RData", "hg38.SCREEN.GRCh38_CTCF.RData")
fileNameOut1.1 <- file.path(dir_project, "RData", "hg38.SCREEN.GRCh38_CTCF.bed")
fileNameOut2 <- file.path(dir_project, "RData", "mm10.SCREEN.mm10_CTCF.RData")
fileNameOut2.1 <- file.path(dir_project, "RData", "mm10.SCREEN.mm10_CTCF.bed")
```

@@

  return(gr)
}
```

```{r}
hg38.SCREEN.GRCh38_CTCF <- mtx_to_gr(mtx = mtx_human, genome_id = "hg38")
length(hg38.SCREEN.GRCh38_CTCF)
summary(width(hg38.SCREEN.GRCh38_CTCF))
mm10.SCREEN.mm10_CTCF <- mtx_to_gr(mtx = mtx_mouse, genome_id = "mm10")
length(mm10.SCREEN.mm10_CTCF)
summary(width(mm10.SCREEN.mm10_CTCF))

# Save as RData object. subdir is the character name of the CTCF GRanges variable
save(list = "hg38.SCREEN.GRCh38_CTCF", file = fileNameOut1)
save(list = "mm10.SCREEN.mm10_CTCF", file = fileNameOut2)
# load(file = paste0(subdir, ".RData"))
# Save BED files
export.bed(hg38.SCREEN.GRCh38_CTCF, fileNameOut1.1)
export.bed(mm10.SCREEN.mm10_CTCF, fileNameOut2.1)
```

### 2022-07-04 EDA_liftOver.Rmd
Latest changes

```{r settings}
# Project folder path
dir_data <- "/Users/mdozmorov/Documents/Data/GoogleDrive/CTCF.dev/CTCF_liftover/"

# Input files
fileNameIn1 <- file.path(dir_data, "sample_annotation.csv")
# Output files
fileNameOut1 <- file.path(dir_data, "Figure_liftOverJaccard.svg")
```

@@

```{r data}
sample_annotation <- read_csv(fileNameIn1)
# Read in data
files <- sample_annotation$File
genome_assembly <- sample_annotation$Assembly
# Object to store GRanges 
bed_list_all <- list() # All chromosomes
bed_list_standard <- list() # Standard chromosomes
# Additional columns
# extraCols_names <- c("character", "character", "character")
# names(extraCols_names) <- c("motif", "pval", "qval")
# Process each file
for(i in 1:length(files)) {
  print(basename(files[i]))
  # Read data
  bed_file <- import(file.path(dir_data, "data", files[i]), format = "bed") # , genome = genome_assembly[i], colnames = c("chrom", "start", "end", "name", "score", "strand", "motif", "pval", "qval"), extraCols = extraCols_names
  # Save all data
  bed_list_all <- c(bed_list_all, list(bed_file))
  # Keep standard chromosomes
  bed_file_standard <- keepStandardChromosomes(bed_file, pruning.mode="coarse")
  # Save filtered data
  bed_list_standard <- c(bed_list_standard, list(bed_file_standard))
}
# Add names
names(bed_list_all) <- sample_annotation$Name
names(bed_list_standard) <- sample_annotation$Name
```

>>>>

```{r settings}
# Project folder path
dir_data <- "/Users/mdozmorov/Documents/Data/GoogleDrive/CTCF.dev/CTCF_liftover/"
dir_project <- "/Users/mdozmorov/Documents/Work/GitHub/CTCF.dev"
# Input files
fileNameIn1 <- file.path(dir_data, "sample_annotation.csv")
# Output files
fileNameOut1 <- file.path(dir_project, "results", "Figure_liftOverJaccard.svg")
```

@@

```{r data}
sample_annotation <- read_csv(fileNameIn1)
# Read in data
files <- sample_annotation$File
genome_assembly <- sample_annotation$Assembly
# Object to store GRanges 
bed_list_all <- list() # All chromosomes
bed_list_standard <- list() # Standard chromosomes
# Additional columns
# extraCols_names <- c("character", "character", "character")
# names(extraCols_names) <- c("motif", "pval", "qval")
# Process each file
for(i in 1:length(files)) {
  print(basename(files[i]))
  # Read data
  # bed_file <- import(file.path(dir_data, "data", files[i]), format = "bed", genome = genome_assembly[i], colnames = c("chrom", "start", "end", "name", "score", "strand", "motif", "pval", "qval"), extraCols = extraCols_names)
  mtx <- read_tsv(file.path(dir_data, "data", files[i]), col_names = FALSE)
  bed_file <- GRanges(seqnames = mtx$X1, ranges = IRanges(start = mtx$X2, end = mtx$X3), strand = mtx$X6)
  # Save all data
  bed_list_all <- c(bed_list_all, list(bed_file))
  # Keep standard chromosomes
  bed_file_standard <- keepStandardChromosomes(bed_file, pruning.mode="coarse")
  # Save filtered data
  bed_list_standard <- c(bed_list_standard, list(bed_file_standard))
}
# Add names
names(bed_list_all) <- sample_annotation$Name
names(bed_list_standard) <- sample_annotation$Name
```

### 2022-07-04 FIMO_processing.Rmd
Latest changes

```{r settings}
# Project folder path
dir_data <- "/Users/mdozmorov/Documents/Data/GoogleDrive/CTCF.dev/merlot"
# Motifs
motif_human  <- c("CIS_BP_2.00_Homo_sapiens", "CTCFBSDB_PWM", "HOCOMOCOv11_core_HUMAN_mono_meme_format", "JASPAR2022_CORE_vertebrates_non_redundant_v2", "jolma2013", "SwissRegulon_human_and_mouse")
motif_mouse  <- c("CIS_BP_2.00_Mus_musculus", "CTCFBSDB_PWM", "HOCOMOCOv11_core_MOUSE_mono_meme_format", "JASPAR2022_CORE_vertebrates_non_redundant_v2", "jolma2013", "SwissRegulon_human_and_mouse")
# Genomes
genome_human <- c("T2T", "hg38", "hg19")
genome_mouse <- c("mm10", "mm9")
# Chromosomes
chromosome_human <- c("chr1", "chr2", "chr3", "chr4", "chr5", "chr6", "chr7", "chr8", "chr9", "chr10", "chr11", "chr12", "chr13", "chr14", "chr15", "chr16", "chr17", "chr18", "chr19", "chr20", "chr21", "chr22", "chrX", "chrY")
chromosome_mouse <- c("chr1", "chr2", "chr3", "chr4", "chr5", "chr6", "chr7", "chr8", "chr9", "chr10", "chr11", "chr12", "chr13", "chr14", "chr15", "chr16", "chr17", "chr18", "chr19", "chrX", "chrY")

# Input files
fileNameIn1 <- file.path(dir_data, "")
# Output files
fileNameOut1 <- file.path(dir_data, "log_PWMs.csv")
```

@@

```{r data}
combine_fimo <- function(motif = "SwissRegulon_human_and_mouse", genome_id = "T2T", chromosomes = chromosome_human) {
  ctcfGR <- c()
  for(chrom in chromosomes) {
    # print(chrom)
    fileNameIn <- file.path(dir_data, paste(motif, genome_id, chrom, sep = "_"), "fimo.txt.gz")
    # Read "fimo.bed" created by "fimo.qsub"
    ctcfBED <- read_tsv(fileNameIn, col_types = c("ccddcdddc"))
    # Convert to GRanges object
    ctcf_chr_gr <- GRanges(seqnames = ctcfBED$`sequence name`, 
                      IRanges(start = ctcfBED$start, end = ctcfBED$stop),
                      strand = ctcfBED$strand)
    # Add metadata
    ctcf_chr_gr$name = ctcfBED$`#pattern name`
    ctcf_chr_gr$score = ctcfBED$score
    ctcf_chr_gr$pvalue = ctcfBED$`p-value`
    ctcf_chr_gr$qvalue = ctcfBED$`q-value`
    ctcf_chr_gr$sequence = ctcfBED$`matched sequence`
    ctcfGR <- c(ctcfGR, ctcf_chr_gr)
  }
  # Combine all chromosomes
  ctcfGR <- do.call(c, as(ctcfGR, "GRangesList"))
  # Sort
  ctcfGR <- ctcfGR %>% sort()

@@

  ctcfGR_to_save <- paste(genome_id, motif, "FIMO", sep = ".")
  fileNameOut1 <- file.path(dir_data, paste0(ctcfGR_to_save, ".RData"))
  fileNameOut1.1 <- file.path(dir_data, paste0(ctcfGR_to_save, ".bed"))
  # Assign this object to the subfolder-specific variable name
  assign(ctcfGR_to_save, ctcfGR)
  # Save as RData object. ctcfGR_to_save is the character name of the CTCF GRanges variable
  save(list = ctcfGR_to_save, file = fileNameOut1)
  # load(file = fileNameOut1)
  # export.bed(eval(parse(text = ctcfGR_to_save)), fileNameOut1.1)
  ctcfDF_to_save <- as.data.frame(eval(parse(text = ctcfGR_to_save)))
  ctcfDF_to_save <- ctcfDF_to_save[, c("seqnames", "start", "end", "name", "score", "strand", "width", "pvalue", "qvalue", "sequence")]
  write_tsv(ctcfDF_to_save, fileNameOut1.1, col_names = FALSE)
}
```

# Human data hg18

```{r hg18}
for (gen in c("hg18", genome_human)) {
  for (mot in "JASPAR2022_MA0139.1") {
    print(paste(gen, mot, sep = "."))
    combine_fimo(motif = mot, genome_id = gen, chromosomes = chromosome_human)
  }
}
```

@@

@@

```{r}
# for file in `find . -type f -name "*MA0139*"`; do
#   echo $file
#   newfile=`echo $file | sed 's/MA0139/JASPAR2022_MA0139/'`
#   echo $newfile
#   mv $file $newfile
# done
# 
files <- list.files(dir_data, pattern = "*.bed", full.names = TRUE)
for (file in files) {
  print(basename(file) %>% str_remove(., ".bed"))
  mtx <- read_tsv(file, col_names = FALSE)
  colnames(mtx) <- c("seqnames", "start", "end", "name", "score", "strand", "width", "pvalue", "qvalue", "sequence")
  gr <- makeGRangesFromDataFrame(mtx)
  # Summary
  summary_to_save <- data.frame(File = basename(file) %>% str_remove(., ".bed"),
                                Assembly = strsplit(basename(file), ".", fixed = TRUE)[[1]][1],
                                All = length(gr),
                                Reduced = length(reduce(gr)))
  write_csv(summary_to_save, fileNameOut1, append = TRUE)
}
```

>>>>

```{r settings}
# Project folder path
dir_data <- "/Users/mdozmorov/Documents/Data/GoogleDrive/CTCF.dev/merlot"
dir_project <- "/Users/mdozmorov/Documents/Work/GitHub/CTCF.dev/"
# Rename motif folders to replace "-" by "_"
# for file in `find . -type d -name "*-*"`; do newfile=`echo $file | sed 's/-/_/g'`; mv $file $newfile; done
# Capitalize Jolma
# for file in `find . -type d -name "jolma*"`; do newfile=`echo $file | sed 's/j/J/'`; mv $file $newfile; done
# Remove "_corrected"
# for file in `find . -type d -name "*_corrected*"`; do newfile=`echo $file | sed 's/_corrected//'`; mv $file $newfile; done
# Motifs
motif_human  <- c("CIS_BP_2.00_Homo_sapiens", "CTCFBSDB_PWM", "HOCOMOCOv11_core_HUMAN_mono_meme_format", "JASPAR2022_CORE_vertebrates_non_redundant_v2", "Jolma2013", "SwissRegulon_human_and_mouse")
motif_mouse  <- c("CIS_BP_2.00_Mus_musculus", "CTCFBSDB_PWM", "HOCOMOCOv11_core_MOUSE_mono_meme_format", "JASPAR2022_CORE_vertebrates_non_redundant_v2", "Jolma2013", "SwissRegulon_human_and_mouse")
# Genomes
genome_human <- c("T2T", "hg38", "hg19")
genome_mouse <- c("mm10", "mm9", "mm39")
# Chromosomes
chromosome_human <- c("chr1", "chr2", "chr3", "chr4", "chr5", "chr6", "chr7", "chr8", "chr9", "chr10", "chr11", "chr12", "chr13", "chr14", "chr15", "chr16", "chr17", "chr18", "chr19", "chr20", "chr21", "chr22", "chrX", "chrY")
chromosome_mouse <- c("chr1", "chr2", "chr3", "chr4", "chr5", "chr6", "chr7", "chr8", "chr9", "chr10", "chr11", "chr12", "chr13", "chr14", "chr15", "chr16", "chr17", "chr18", "chr19", "chrX", "chrY")

# Input files
fileNameIn1 <- file.path(dir_data, "")
```

@@

```{r data}
# Function to combine database- and chromosome-specific FIMO results for a given genome assembly
combine_fimo <- function(motif = "SwissRegulon_human_and_mouse", genome_id = "T2T", chromosomes = chromosome_human) {
  ctcfGR <- c()
  for(chrom in chromosomes) {
    # print(chrom)
    fileNameIn <- file.path(dir_data, "results", paste(motif, genome_id, chrom, sep = "_"), "fimo.txt.gz")
    # Read "fimo.bed" created by "fimo.qsub"
    ctcfBED <- read_tsv(fileNameIn, col_types = c("ccddcdddc"))
    # Convert to GRanges object
    ctcf_chr_gr <- GRanges(seqnames = ctcfBED$`sequence name`, 
                      IRanges(start = ctcfBED$start, end = ctcfBED$stop),
                      strand = ctcfBED$strand)
    # Add metadata
    ctcf_chr_gr$name = ctcfBED$`#pattern name`
    ctcf_chr_gr$score = ctcfBED$score
    ctcf_chr_gr$pvalue = ctcfBED$`p-value`
    ctcf_chr_gr$qvalue = ctcfBED$`q-value`
    ctcf_chr_gr$sequence = ctcfBED$`matched sequence`
    ctcfGR <- c(ctcfGR, ctcf_chr_gr)
  }
  # Combine all chromosomes
  ctcfGR <- do.call(c, as(ctcfGR, "GRangesList"))
  # Sort
  ctcfGR <- ctcfGR %>% sort()

@@

  ctcfGR_to_save <- paste(genome_id, motif, sep = ".")
  fileNameOut1 <- file.path(dir_project, "RData", paste0(ctcfGR_to_save, ".RData"))
  fileNameOut1.1 <- file.path(dir_project, "RData", paste0(ctcfGR_to_save, ".bed"))
  # Assign this object to the subfolder-specific variable name
  assign(ctcfGR_to_save, ctcfGR)
  # Save as RData object. ctcfGR_to_save is the character name of the CTCF GRanges variable
  save(list = ctcfGR_to_save, file = fileNameOut1)
  # load(file = fileNameOut1)
  # export.bed(eval(parse(text = ctcfGR_to_save)), fileNameOut1.1)
  ctcfDF_to_save <- as.data.frame(eval(parse(text = ctcfGR_to_save)))
  ctcfDF_to_save <- ctcfDF_to_save[, c("seqnames", "start", "end", "name", "score", "strand", "width", "pvalue", "qvalue", "sequence")]
  write_tsv(ctcfDF_to_save, fileNameOut1.1, col_names = FALSE)
}
```

# Human data for liftOver

```{r hg}
for (gen in c("hg18", genome_human)) {
  for (mot in "MA0139.1") {
    print(paste(gen, mot, sep = "."))
    combine_fimo(motif = mot, genome_id = gen, chromosomes = chromosome_human)
  }
}
```

@@

# Mouse data for liftOver

```{r mm}
for (gen in c(genome_mouse)) {
  for (mot in "MA0139.1") {
    print(paste(gen, mot, sep = "."))
    combine_fimo(motif = mot, genome_id = gen, chromosomes = chromosome_mouse)
  }
}
```

@@

```{r}
# for file in `find . -type f -name "*MA0139*"`; do
#   echo $file
#   newfile=`echo $file | sed 's/MA0139/JASPAR2022_MA0139/'`
#   echo $newfile
#   mv $file $newfile
# done
# 
# Output files
fileNameOut1 <- file.path(dir_project, "results", "log_PWMs.csv")
if (file.exists(fileNameOut1)) { file.remove(fileNameOut1) }
# Defined from FIMO_EDA
pval_threshold <- 1e-6
files <- list.files(file.path(dir_project, "RData"), pattern = "bed", full.names = TRUE)
for (fil in files) {
  print(basename(fil) %>% str_remove(., ".bed"))
  mtx <- read_tsv(fil, col_names = FALSE, col_types = c("ciicdciddc"))
  colnames(mtx) <- c("seqnames", "start", "end", "name", "score", "strand", "width", "pvalue", "qvalue", "sequence")
  gr <- makeGRangesFromDataFrame(mtx, keep.extra.columns = TRUE)
  # Summary
  summary_to_save <- data.frame(File = basename(fil) %>% str_remove(., ".bed"),
                                Assembly = strsplit(basename(fil), ".", fixed = TRUE)[[1]][1],
                                `All (p-value threshold 1e-4)` = length(gr),
                                `Reduced (p-value threshold 1e-4)` = length(reduce(gr)),
                                `All (p-value threshold 1e-6)` = length(gr[gr$pvalue < pval_threshold]),
                                `Reduced (p-value threshold 1e-6)` = length(reduce(gr[gr$pvalue < pval_threshold])) )
  write_csv(summary_to_save, fileNameOut1, append = TRUE)
  gc()
}
```

### 2022-07-04 manuscript/manuscript_bioinformatics.Rmd
Latest changes

Position Weight Matrices (PWMs) [@Stormo:2000uw] of CTCF sequence motifs from Jaspar 2022 [@Castro-Mondragon:2022uv], HOCOMOCO v11 [@Kulakovskiy:2018vw], CIS-BP [@Weirauch:2014ul], Jolma 2013 [@Jolma:2013vd], and SwissRegulon [@Pachkov:2013uk] databases were downloaded from the MEME Motif database (`Table S1`). They were largely similar with the exception of Jolma 2013 and two long (34bp and 35bp) Jaspar PWMs (`Figure S1A`). The CTCF binding site database (CTCFBSDB) [@Ziebarth:2013vj] includes PWMs for the M1+M2 motifs [@Schmidt:2012uk], the Ren_20 motif [@Kim:2007ux], and the LM2, LM7, and LM23 motifs [@Xie:2007wm] (`Figure S1B`). The CIS-BP database [@Weirauch:2014ul] lists 83 human-specific and two mouse-specific PWMs (`Figure S1CD`). These PWMs were used to scan the human and mouse genomes for CTCF binding motifs using the FIMO tool from the MEME suite [@Bailey:2009aa].

>>>>

Position Weight Matrices (PWMs) [@Stormo:2000uw] of CTCF sequence motifs from Jaspar 2022 [@Castro-Mondragon:2022uv], HOCOMOCO v11 [@Kulakovskiy:2018vw], Jolma 2013 [@Jolma:2013vd], and SwissRegulon [@Pachkov:2013uk] databases were downloaded from the MEME Motif database (`Table S1`). They were largely similar with the exception of Jolma 2013 and two long (34bp and 35bp) Jaspar PWMs (`Figure S1A`). The CTCF binding site database (CTCFBSDB) [@Ziebarth:2013vj] includes PWMs for the M1+M2 motifs [@Schmidt:2012uk], the Ren_20 motif [@Kim:2007ux], and the LM2, LM7, and LM23 motifs [@Xie:2007wm] (`Figure S1B`). The CIS-BP database [@Weirauch:2014ul] lists 83 human-specific and two mouse-specific PWMs (`Figure S1CD`). These PWMs were used to scan the human and mouse genomes for CTCF binding motifs using the FIMO tool from the MEME suite [@Bailey:2009aa].
